\chapter{Grid and Laplacian}
\label{chap:grid-and-laplacian}

\begin{quote}
Spec dir: \passthrough{\lstinline!specs/2026-05-01-phase-1-laplacian/!}.
Version: \passthrough{\lstinline!0.1.0!}. \emph{Note: this note was
written retroactively after Phase 4, so the ``current PR'' framing of
later notes is absent here.}
\end{quote}

\subsection{What this gives you}\label{what-this-gives-you}

The structural foundation of the library: a Cartesian-orthogonal
\textbf{sampling grid}, a \textbf{periodic scalar field} built on it,
and a \textbf{2nd-order Laplacian} operator. Everything later in the
roadmap (gradient, divergence, integration, functional evaluation,
\ldots) sits on top of these three pieces.

\subsection{API}\label{api}

Headers:
\href{../../../include/gridcalc/core/Grid.hpp}{\passthrough{\lstinline!include/gridcalc/core/Grid.hpp!}},
\href{../../../include/gridcalc/core/Field.hpp}{\passthrough{\lstinline!include/gridcalc/core/Field.hpp!}},
\href{../../../include/gridcalc/core/IndexPolicy.hpp}{\passthrough{\lstinline!include/gridcalc/core/IndexPolicy.hpp!}},
\href{../../../include/gridcalc/stencil/CentralDifference.hpp}{\passthrough{\lstinline!include/gridcalc/stencil/CentralDifference.hpp!}},
\href{../../../include/gridcalc/diff/Laplacian.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Laplacian.hpp!}}.

\begin{lstlisting}[language={C++}]
namespace gridcalc::core {

class Grid {
 public:
  Grid(int Nx, int Ny, int Nz, const Vec3d& cell_size);
  int          getNx() const noexcept;
  int          getNy() const noexcept;
  int          getNz() const noexcept;
  const Vec3d& getCellSize() const noexcept;
  double       getCellVolume() const noexcept;
  std::size_t  getSize() const noexcept;
  std::size_t  getLinearIndex(int i, int j, int k) const noexcept;
};

template <class T, class Policy = IndexPolicy::Periodic>
class Field {
 public:
  explicit Field(const Grid& grid);                // zero-initialized
  Field(const Grid& grid, const T& value);         // broadcast
  template <class F> Field(const Grid& grid, F&& f);  // sample f(x, y, z)

  T&       operator()(int i, int j, int k) noexcept;     // wraps via Policy
  const T& operator()(int i, int j, int k) const noexcept;

  const Grid&  getGrid() const noexcept;
  std::size_t  getSize() const noexcept;
  T*           data() noexcept;
  const T*     data() const noexcept;
  T*           begin() noexcept;  T* end() noexcept;
};

namespace IndexPolicy {
  struct Periodic { static constexpr int wrap(int i, int N) noexcept; };
}

}  // namespace gridcalc::core

namespace gridcalc::diff {
core::Field<double> laplacian(const core::Field<double>& field);
}
\end{lstlisting}

\subsubsection{Storage layout}\label{storage-layout}

\passthrough{\lstinline!Field<T>!} storage is \textbf{i-fastest}: the
linear offset for \passthrough{\lstinline!(i, j, k)!} is
\passthrough{\lstinline!i + Nx * (j + Ny * k)!}. This matches the Eigen
/ FFT convention and is part of the public contract from Phase 1 forward
--- Phase 9 (spectral verification) and Phase 20 (benchmarks) both rely
on it.

\subsubsection{Index policy}\label{index-policy}

\passthrough{\lstinline!Field<T, Policy>!} defaults to
\passthrough{\lstinline!IndexPolicy::Periodic!}, which wraps
out-of-range indices modulo \passthrough{\lstinline!N!}.
\passthrough{\lstinline!field(-1, j, k)!} returns
\passthrough{\lstinline!field(Nx - 1, j, k)!}.
\passthrough{\lstinline!Dirichlet!} and
\passthrough{\lstinline!Neumann!} policies arrive in Phase 18; the
policy template parameter is the only piece of forward-looking design
admitted at Phase 1.

\subsection{Worked example}\label{worked-example}

Recover the Laplacian eigenvalue of a trig product:

\begin{lstlisting}[language={C++}]
#include <cmath>
#include <gridcalc/core/Field.hpp>
#include <gridcalc/core/Grid.hpp>
#include <gridcalc/diff/Laplacian.hpp>

using gridcalc::core::Field;
using gridcalc::core::Grid;
using gridcalc::core::Vec3d;
using gridcalc::diff::laplacian;

constexpr double kPi = 3.14159265358979323846;

const int N = 32;
const double L = 2.0 * kPi;
const double h = L / N;

Grid grid(N, N, N, Vec3d(h, h, h));
Field<double> psi(grid, [](double x, double y, double z) {
  return std::sin(x) * std::sin(y) * std::sin(z);
});

Field<double> lap = laplacian(psi);
// lap ~= -3 * psi  (since the analytical Laplacian eigenvalue is -3)
// Relative max-norm error ~ k^4 h^2 / (12 k^2) ~ h^2 / 12 ~ 3e-3 at N = 32.
\end{lstlisting}

\subsection{\texorpdfstring{What this does \emph{not} do
(yet)}{What this does not do (yet)}}\label{what-this-does-not-do-yet}

\begin{itemize}
\tightlist
\item
  \textbf{Periodic boundary only.} Dirichlet / Neumann policies arrive
  in Phase 18.
\item
  \textbf{2nd-order accuracy only.} Higher-order accuracy
  (\passthrough{\lstinline!Order = 4!}) on the same operator arrives in
  Phase 7.
\item
  \textbf{Scalar \passthrough{\lstinline!T = double!} only in practice.}
  Although \passthrough{\lstinline!Field<T>!} is templated, only
  \passthrough{\lstinline!double!} is exercised at Phase 1.
  \passthrough{\lstinline!Field<Vec3d>!} arrives in Phase 2's gradient
  output.
\end{itemize}

For the why-this-stencil-converges background --- derivation, error
analysis, references --- see
\href{../../developer-note/notes/phase-1-laplacian.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-1-laplacian.md!}}.
