\chapter{Gradient and divergence}
\label{chap:gradient-and-divergence}

\begin{quote}
Spec dir:
\passthrough{\lstinline!specs/2026-05-02-phase-2-gradient-divergence/!}.
Version: \passthrough{\lstinline!0.2.0!}. \emph{Note: this note was
written retroactively after Phase 4 (see specs/STATUS.md ``Decisions
worth knowing'').}
\end{quote}

\subsection{What this gives you}\label{what-this-gives-you}

The basic first-order vector operators on a periodic scalar field:

\begin{itemize}
\tightlist
\item
  \textbf{Gradient:}
  \passthrough{\lstinline!gradient(ψ) -> Field<Vec3d>!}. At each grid
  point, returns the Cartesian triple
  \((\partial_x \psi,\, \partial_y \psi,\, \partial_z \psi)\).
\item
  \textbf{Divergence:}
  \passthrough{\lstinline!divergence(V) -> Field<double>!}. For a vector
  field \(\mathbf{V}\) stored as \passthrough{\lstinline!Field<Vec3d>!},
  returns \(\partial_x V_x + \partial_y V_y + \partial_z V_z\) at each
  grid point.
\end{itemize}

Both are 2nd-order accurate central differences applied per axis, with
periodic wrap delegated to the input \passthrough{\lstinline!Field!}'s
\passthrough{\lstinline!Policy!} (\passthrough{\lstinline!Periodic!}
from Phase 1 onward).

\subsection{API}\label{api}

Headers:
\href{../../../include/gridcalc/diff/Gradient.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Gradient.hpp!}},
\href{../../../include/gridcalc/diff/Divergence.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Divergence.hpp!}},
\href{../../../include/gridcalc/stencil/FirstDerivative.hpp}{\passthrough{\lstinline!include/gridcalc/stencil/FirstDerivative.hpp!}},
\href{../../../include/gridcalc/core/EigenAliases.hpp}{\passthrough{\lstinline!include/gridcalc/core/EigenAliases.hpp!}}.

\begin{lstlisting}[language={C++}]
namespace gridcalc::core {
using Vec3d = Eigen::Vector3d;
}

namespace gridcalc::diff {

core::Field<core::Vec3d> gradient   (const core::Field<double>&)         noexcept;
core::Field<double>      divergence (const core::Field<core::Vec3d>&)    noexcept;

}  // namespace gridcalc::diff
\end{lstlisting}

The \passthrough{\lstinline!Vec3d!} alias was relocated from
\passthrough{\lstinline!core/Grid.hpp!} (Phase 1) to
\passthrough{\lstinline!core/EigenAliases.hpp!} in Phase 2 once a second
operator needed it. The fully-qualified name
\passthrough{\lstinline!gridcalc::core::Vec3d!} is unchanged from Phase
1.

\subsection{Worked example}\label{worked-example}

Compute a gradient, then a divergence, then verify that
\(\nabla \cdot \nabla \psi\) converges to \(\nabla^2 \psi\) at order 2:

\begin{lstlisting}[language={C++}]
#include <cmath>
#include <gridcalc/core/EigenAliases.hpp>
#include <gridcalc/core/Field.hpp>
#include <gridcalc/core/Grid.hpp>
#include <gridcalc/diff/Divergence.hpp>
#include <gridcalc/diff/Gradient.hpp>
#include <gridcalc/diff/Laplacian.hpp>

using gridcalc::core::Field;
using gridcalc::core::Grid;
using gridcalc::core::Vec3d;
using gridcalc::diff::divergence;
using gridcalc::diff::gradient;
using gridcalc::diff::laplacian;

constexpr double kPi = 3.14159265358979323846;

const int N = 32;
const double L = 2.0 * kPi;
const double h = L / N;

Grid grid(N, N, N, Vec3d(h, h, h));
Field<double> psi(grid, [](double x, double y, double z) {
  return std::sin(x) * std::sin(y) * std::sin(z);
});

Field<Vec3d>  grad_psi  = gradient(psi);
Field<double> lap_psi   = laplacian(psi);
Field<double> indirect  = divergence(grad_psi);
// indirect ~ lap_psi at order 2 in h, but the two are *not* identically
// equal at finite h -- divergence(gradient(psi)) is a stride-2 stencil
// while laplacian(psi) is a 3-point stencil. They share the same
// continuum limit but have different leading-order error constants.
\end{lstlisting}

\subsection{Vector-field example for
divergence}\label{vector-field-example-for-divergence}

\passthrough{\lstinline!divergence!} takes any
\passthrough{\lstinline!Field<Vec3d>!}, not only one produced by
\passthrough{\lstinline!gradient!}. For
\(\mathbf{V} = (\sin x \cos y,\, \cos x \sin y,\, \sin z)\) on
\([0, 2\pi]^3\), the analytical divergence is
\(2\cos x \cos y + \cos z\):

\begin{lstlisting}[language={C++}]
Field<Vec3d> V(grid, [](double x, double y, double z) {
  return Vec3d(std::sin(x) * std::cos(y),
               std::cos(x) * std::sin(y),
               std::sin(z));
});
Field<double> div_V = divergence(V);
// div_V ~ 2 cos(x) cos(y) + cos(z), with relative max-norm error O(h^2).
\end{lstlisting}

\subsection{\texorpdfstring{What this does \emph{not} do
(yet)}{What this does not do (yet)}}\label{what-this-does-not-do-yet}

\begin{itemize}
\tightlist
\item
  \textbf{No curl operator.} Vector calculus on
  \passthrough{\lstinline!Field<Vec3d>!} beyond divergence arrives in
  later phases.
\item
  \textbf{No 4th-order accuracy.}
  \passthrough{\lstinline!stencil::FirstDerivative<4>!} is deferred to
  Phase 7 alongside \passthrough{\lstinline!Coefficients<4>!}.
\item
  \textbf{Only periodic wrap.} Dirichlet / Neumann gradients arrive in
  Phase 18.
\item
  \textbf{No per-component-decomposition helper.}
  \passthrough{\lstinline!Field<Vec3d>!} is the canonical gradient
  return; iterate over it and project per axis if you need a
  \passthrough{\lstinline!Field<double>!} for one component.
\end{itemize}

For the why-divergence(gradient)-isn't-laplacian background and the full
error analysis --- see
\href{../../developer-note/notes/phase-2-gradient-divergence.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-2-gradient-divergence.md!}}.
