\chapter{Explicit-Euler diffusion}
\label{chap:explicit-euler-diffusion}

\begin{quote}
Spec dir:
\passthrough{\lstinline!specs/2026-05-02-phase-5-explicit-euler/!}.
Version: \passthrough{\lstinline!0.5.0!}.
\end{quote}

\subsection{What this gives you}\label{what-this-gives-you}

Time integration enters the picture. Two new public functions in the
\passthrough{\lstinline!gridcalc::solve!} namespace:

\begin{itemize}
\tightlist
\item
  \textbf{\passthrough{\lstinline!explicitEuler!}} --- a generic
  forward-Euler integrator over an arbitrary RHS callable.
\item
  \textbf{\passthrough{\lstinline!diffuse!}} --- a thin convenience
  driver for the heat equation \(\partial_t \psi = D \nabla^2 \psi\),
  with a built-in CFL stability check.
\end{itemize}

Both mutate the input \passthrough{\lstinline!Field<double>!} \textbf{in
place} --- the natural pattern for iterative time stepping. If you want
to keep the initial condition, copy the field first.

\subsection{API}\label{api}

Headers:
\href{../../../include/gridcalc/solve/ExplicitEuler.hpp}{\passthrough{\lstinline!include/gridcalc/solve/ExplicitEuler.hpp!}},
\href{../../../include/gridcalc/solve/Diffusion.hpp}{\passthrough{\lstinline!include/gridcalc/solve/Diffusion.hpp!}}.

\begin{lstlisting}[language={C++}]
namespace gridcalc::solve {

template <class Rhs>
void explicitEuler(core::Field<double>& psi,
                   Rhs&& rhs,
                   double dt,
                   int n_steps);

void diffuse(core::Field<double>& psi,
             double D,
             double dt,
             int n_steps);

}  // namespace gridcalc::solve
\end{lstlisting}

\subsubsection{\texorpdfstring{\texttt{explicitEuler}
semantics}{explicitEuler semantics}}\label{expliciteuler-semantics}

At each step:
\passthrough{\lstinline!tmp = rhs(psi); psi[n] += dt * tmp[n]!}. The RHS
callable must be invocable as
\passthrough{\lstinline!Field<double> rhs(const Field<double>\&)!};
mismatch trips a clean \passthrough{\lstinline!static\_assert!} at the
call site.

The integrator does \textbf{not} validate stability --- that depends on
the specific RHS, which is invisible to the integrator. If your RHS has
a known stability bound, check it yourself before calling.

Throws \passthrough{\lstinline!std::invalid\_argument!} on
\passthrough{\lstinline!dt < 0!} or
\passthrough{\lstinline!n\_steps < 0!}.

\subsubsection{\texorpdfstring{\texttt{diffuse}
semantics}{diffuse semantics}}\label{diffuse-semantics}

Advances the heat equation by \passthrough{\lstinline!n\_steps!}
explicit-Euler steps of size \passthrough{\lstinline!dt!}. Internally
calls \passthrough{\lstinline!explicitEuler!} with
\passthrough{\lstinline!diff::laplacian!} as the RHS and
\passthrough{\lstinline!D * dt!} as the effective step (mathematically
identical to running an unscaled Laplacian RHS at a scaled time step;
the trick avoids a per-step scaling loop).

\textbf{CFL stability check.} Before integration begins,
\passthrough{\lstinline!diffuse!} validates the von Neumann bound

\[
D \cdot dt \cdot \left( \frac{1}{h_x^2} + \frac{1}{h_y^2} + \frac{1}{h_z^2} \right) \;\le\; \frac{1}{2}
\]

and throws \passthrough{\lstinline!std::invalid\_argument!} (with a
message naming the offending values) if the bound is exceeded. For
isotropic spacing this reduces to \(D\,dt/h^2 \le 1/(2d) = 1/6\) in 3D,
the form quoted in the roadmap.

Throws on negative \passthrough{\lstinline!D!},
\passthrough{\lstinline!dt!}, or \passthrough{\lstinline!n\_steps!}.

\subsection{Worked example}\label{worked-example}

Diffuse a trig eigenfunction and compare to the analytical decay:

\begin{lstlisting}[language={C++}]
#include <cmath>
#include <gridcalc/core/Field.hpp>
#include <gridcalc/core/Grid.hpp>
#include <gridcalc/solve/Diffusion.hpp>

using gridcalc::core::Field;
using gridcalc::core::Grid;
using gridcalc::core::Vec3d;
using gridcalc::solve::diffuse;

constexpr double kPi = 3.14159265358979323846;

const int N = 32;
const double L = 2.0 * kPi;
const double h = L / N;
const double D = 0.1;
const double dt = 0.05;
const int n_steps = 100;
const double T = n_steps * dt;  // 5.0

Grid grid(N, N, N, Vec3d(h, h, h));

// Initial condition: psi_0 = sin(x) sin(y) sin(z)
// is an eigenfunction of the Laplacian with eigenvalue -3.
Field<double> psi(grid, [](double x, double y, double z) {
  return std::sin(x) * std::sin(y) * std::sin(z);
});

// In place: psi is mutated.
diffuse(psi, D, dt, n_steps);

// After integration, psi ~= exp(-3 D T) * sin(x) sin(y) sin(z),
// to within stencil-order error (~few percent at N=32).
\end{lstlisting}

\subsection{\texorpdfstring{Custom RHS via
\texttt{explicitEuler}}{Custom RHS via explicitEuler}}\label{custom-rhs-via-expliciteuler}

If you have an RHS that isn't \(D\nabla^2\psi\), call
\passthrough{\lstinline!explicitEuler!} directly and provide your own
callable. \textbf{You} are responsible for the relevant stability check.

\begin{lstlisting}[language={C++}]
#include <gridcalc/diff/Laplacian.hpp>
#include <gridcalc/solve/ExplicitEuler.hpp>

// Reaction-diffusion: d_t psi = D laplacian(psi) - k * psi
using gridcalc::diff::laplacian;
using gridcalc::solve::explicitEuler;

const double D = 0.1;
const double k = 0.05;

auto rhs = [D, k](const Field<double>& f) {
  Field<double> out = laplacian(f);
  // out *= D; subtract k * f
  const std::size_t N = out.getSize();
  double* o = out.data();
  const double* p = f.data();
  for (std::size_t n = 0; n < N; ++n) {
    o[n] = D * o[n] - k * p[n];
  }
  return out;
};

explicitEuler(psi, rhs, dt, n_steps);
\end{lstlisting}

\subsection{\texorpdfstring{What this does \emph{not} do
(yet)}{What this does not do (yet)}}\label{what-this-does-not-do-yet}

\begin{itemize}
\tightlist
\item
  \textbf{No higher-order time integrators.} RK4 + a generic
  \passthrough{\lstinline!solve::Integrator!} interface arrive in Phase
  6; the explicit-Euler entry point will be refactored to consume that
  interface then.
\item
  \textbf{No adaptive time stepping.}
\item
  \textbf{No implicit / Crank--Nicolson diffusion.} Phase 19 introduces
  implicit diffusion, after Phase 18 lands non-periodic boundary
  policies.
\item
  \textbf{No automatic CFL adjustment.}
  \passthrough{\lstinline!diffuse!} throws; the caller picks a smaller
  \passthrough{\lstinline!dt!} and retries.
\item
  \textbf{No threading.} Phase 20 (performance pass) parallelizes the
  per-step accumulation loop.
\end{itemize}

For the why-this-stability-bound and the time-discretization error
analysis, see
\href{../../developer-note/notes/phase-5-explicit-euler.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-5-explicit-euler.md!}}.
