\chapter{Higher-order accuracy stencils}
\label{chap:higher-order-accuracy-stencils}

\begin{quote}
Spec dir:
\passthrough{\lstinline!specs/2026-05-03-phase-7-higher-order-stencils/!}.
Version: \passthrough{\lstinline!0.7.0!}.
\end{quote}

\subsection{What this gives you}\label{what-this-gives-you}

A compile-time switch on the three Phase 1--2 differential operators
that selects the \textbf{stencil accuracy order}. The order-2 stencils
that have shipped since Phase 1/2 stay the default; opting into order 4
gives an \passthrough{\lstinline!O(h\^4)!} truncation error instead of
\passthrough{\lstinline!O(h\^2)!} --- typically \textbf{10× to 100×
smaller absolute error at the same grid resolution}.

\begin{lstlisting}[language={C++}]
diff::laplacian(psi);          // order 2 (Phase 1 default; unchanged)
diff::laplacian<4>(psi);       // order 4 (new)

diff::gradient(psi);           // order 2 (Phase 2 default; unchanged)
diff::gradient<4>(psi);        // order 4

diff::divergence(V);           // order 2
diff::divergence<4>(V);        // order 4
\end{lstlisting}

All Phase 1/2/3/4/5/6 callers keep working unchanged because the new
template parameter has a default.

\subsection{API}\label{api}

Headers (unchanged file paths; signatures gain a template parameter):
\href{../../../include/gridcalc/diff/Laplacian.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Laplacian.hpp!}},
\href{../../../include/gridcalc/diff/Gradient.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Gradient.hpp!}},
\href{../../../include/gridcalc/diff/Divergence.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Divergence.hpp!}}.

\begin{lstlisting}[language={C++}]
namespace gridcalc::diff {

template <int Order = 2>
core::Field<double>      laplacian (const core::Field<double>&)        noexcept;

template <int Order = 2>
core::Field<core::Vec3d> gradient  (const core::Field<double>&)        noexcept;

template <int Order = 2>
core::Field<double>      divergence(const core::Field<core::Vec3d>&)   noexcept;

}  // namespace gridcalc::diff
\end{lstlisting}

\passthrough{\lstinline!Order!} must be a supported even integer. Phase
1 specialized the order-2 stencils; Phase 7 specializes the order-4
stencils. Calling e.g.~\passthrough{\lstinline!laplacian<3>(psi)!}
triggers a \textbf{compile error} at the call site (the primary
\passthrough{\lstinline!Coefficients<Order>!} template is intentionally
undefined), not a silent zero-stencil bug.

\subsubsection{What ``order 4'' buys you}\label{what-order-4-buys-you}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3750}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3068}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3182}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Property
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
order 2 (default)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
order 4
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Stencil radius & \passthrough{\lstinline!1!} &
\passthrough{\lstinline!2!} \\
Samples read per axis & \passthrough{\lstinline!3!} &
\passthrough{\lstinline!5!} \\
Truncation error & \passthrough{\lstinline!O(h\^2)!} &
\passthrough{\lstinline!O(h\^4)!} \\
Per-point arithmetic cost & \textasciitilde baseline &
\textasciitilde{}\passthrough{\lstinline!5/3 ≈ 1.67×!} baseline \\
Practical absolute error at N=32, ψ = sin·sin·sin &
\textasciitilde{}\passthrough{\lstinline!1e-2!} &
\textasciitilde{}\passthrough{\lstinline!5e-4!} \\
\end{longtable}
}

Order 4 is typically the right pick when: - You want better accuracy at
the same \passthrough{\lstinline!N!} and don't want to refine the grid
(which would be 8× the memory in 3D). - You're computing functionals
where the gradient or Laplacian dominates the error budget (Phase 4 GL
example, Phase 5 / 6 diffusion).

Stay on order 2 when: - The integrand is non-smooth and 4th-order
accuracy doesn't help (the truncation analysis assumes enough
smoothness). - Per-point cost matters and order 2 is ``good enough''. -
You're benchmarking against existing Phase 1/2 results that used order
2.

\subsection{Worked example}\label{worked-example}

Direct comparison on the trig eigenfunction at fixed N:

\begin{lstlisting}[language={C++}]
#include <cmath>
#include <gridcalc/core/EigenAliases.hpp>
#include <gridcalc/core/Field.hpp>
#include <gridcalc/core/Grid.hpp>
#include <gridcalc/diff/Laplacian.hpp>

using gridcalc::core::Field;
using gridcalc::core::Grid;
using gridcalc::core::Vec3d;
using gridcalc::diff::laplacian;

constexpr double kPi = 3.14159265358979323846;

const int N = 32;
const double L = 2.0 * kPi;
const double h = L / N;

Grid grid(N, N, N, Vec3d(h, h, h));
Field<double> psi(grid, [](double x, double y, double z) {
  return std::sin(x) * std::sin(y) * std::sin(z);
});
Field<double> expected(grid, [](double x, double y, double z) {
  return -3.0 * std::sin(x) * std::sin(y) * std::sin(z);
});

Field<double> lap_o2 = laplacian<2>(psi);   // ~1% relative error at N=32
Field<double> lap_o4 = laplacian<4>(psi);   // ~0.05% relative error at N=32
\end{lstlisting}

The 20× absolute-accuracy improvement comes for free at the call site
--- just change the template arg.

\subsection{Mixing with Phase 5/6 time
integration}\label{mixing-with-phase-56-time-integration}

\passthrough{\lstinline!solve::diffuse(psi, D, dt, n\_steps)!} calls
\passthrough{\lstinline!diff::laplacian(...)!} (order 2) under the hood.
To switch the diffusion driver to a 4th-order spatial Laplacian, you
currently need to bypass \passthrough{\lstinline!diffuse!} and use the
generic integrator directly:

\begin{lstlisting}[language={C++}]
#include <gridcalc/solve/Integrator.hpp>

const double D = 0.1;
solve::integrate(
    psi,
    [D](const Field<double>& f) {
      auto out = laplacian<4>(f);          // order-4 Laplacian
      const std::size_t N = out.getSize();
      double* o = out.data();
      for (std::size_t n = 0; n < N; ++n) o[n] *= D;
      return out;
    },
    dt, n_steps,
    solve::RK4{});
\end{lstlisting}

\passthrough{\lstinline!solve::diffuse!} will gain an order template
parameter in a future phase if the use case justifies it. Until then,
the generic-integrator path above is the escape hatch.

\subsection{\texorpdfstring{What this does \emph{not} do
(yet)}{What this does not do (yet)}}\label{what-this-does-not-do-yet}

\begin{itemize}
\tightlist
\item
  \textbf{Order ≥ 6.} Out of scope until a phase actually needs them.
\item
  \textbf{Order 3 (asymmetric / upwind stencils).} Symmetric central
  differences have even truncation orders only.
\item
  \textbf{Higher-order spatial derivatives}
  (\passthrough{\lstinline!∂³!}, \passthrough{\lstinline!∂⁴!}, mixed
  partials). Phase 8.
\item
  \textbf{Order parameter on \passthrough{\lstinline!solve::diffuse!}
  itself.} See workaround above.
\item
  \textbf{Non-orthogonal Bravais grids.} Phase 15. Cross-derivative
  terms appear and the per-axis-independent stencil structure breaks
  down.
\end{itemize}

For the Taylor-matching derivation of the order-4 weights and the
truncation-error analysis, see
\href{../../developer-note/notes/phase-7-higher-order-stencils.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-7-higher-order-stencils.md!}}.
