\chapter{Spectral verification}
\label{chap:spectral-verification}

\begin{quote}
Spec dir:
\passthrough{\lstinline!specs/2026-05-03-phase-9-spectral-verification/!}.
Version: \passthrough{\lstinline!0.9.0!}.
\end{quote}

\subsection{What this gives you}\label{what-this-gives-you}

A second, independent path for computing periodic-grid derivatives via
FFT. \textbf{Not a production differentiation engine.} The
finite-difference operators in \passthrough{\lstinline!gridcalc::diff!}
(Phases 1, 2, 7, 8) are the production code; the
\passthrough{\lstinline!gridcalc::spectral!} namespace exists to be the
gold-standard reference that gates FD changes in CI. Use the spectral
path when you want a verification check, not when you're solving a real
problem.

\begin{lstlisting}[language={C++}]
diff::laplacian<4>(psi);        // production differentiation
spectral::laplacian(psi);       // verification reference (FFT-based)
\end{lstlisting}

The library ships a parameterized GoogleTest fixture
(\href{../../../test/fd_fft_crosscheck_test.cpp}{\passthrough{\lstinline!test/fd\_fft\_crosscheck\_test.cpp!}})
that already cross-checks every Phase 1--8 FD operator at every
advertised accuracy order against its spectral counterpart on every CI
run. Adding a new FD operator in a future phase means adding one more
line to that fixture; the gate catches accidental accuracy regressions
automatically.

\subsection{API}\label{api}

Headers under
\href{../../../include/gridcalc/spectral/}{\passthrough{\lstinline!include/gridcalc/spectral/!}}:

\begin{lstlisting}[language={C++}]
namespace gridcalc::spectral {

// Generic multi-index spectral partial. Multiplies the spectrum by
// (i kx)^Nx (i ky)^Ny (i kz)^Nz; zeros the per-axis Nyquist mode for
// any axis with an odd derivative count.
template <int Nx, int Ny, int Nz>
core::Field<double> partial(const core::Field<double>& field);

// Fused single-FFT-pair operators for the contracted / rank-1 cases.
core::Field<double>      laplacian (const core::Field<double>& field);   // -|k|^2
core::Field<double>      biharmonic(const core::Field<double>& field);   //  |k|^4
core::Field<core::Vec3d> gradient  (const core::Field<double>& field);   // i k

// Lower-level FFT primitives (rarely needed directly).
ComplexField forwardR2C  (const core::Field<double>& field);
core::Field<double>
             backwardC2R (const ComplexField& spectrum, const core::Grid& grid);

// Wavenumber helpers.
std::vector<double> kxRfft(const core::Grid& grid);  // length Nx/2 + 1
std::vector<double> kyFull(const core::Grid& grid);  // length Ny (signed)
std::vector<double> kzFull(const core::Grid& grid);  // length Nz (signed)

}  // namespace gridcalc::spectral
\end{lstlisting}

The Phase 8 d-prefix partials (\passthrough{\lstinline!dxx!},
\passthrough{\lstinline!d3dx2dy!}, \passthrough{\lstinline!d4dx2dy2!},
etc.) do not have one-for-one named
\passthrough{\lstinline!spectral::*!} counterparts. Use
\passthrough{\lstinline!spectral::partial!} with the matching
multi-index instead --- for example, the spectral analog of
\passthrough{\lstinline!diff::d3dx2dy<4>(psi)!} is
\passthrough{\lstinline!spectral::partial<2, 1, 0>(psi)!}.

\subsection{Build flag}\label{build-flag}

\begin{lstlisting}
cmake -B build -DGRIDCALC_USE_FFT=ON   # default: builds the spectral harness
cmake -B build -DGRIDCALC_USE_FFT=OFF  # skips spectral headers and CI gate
\end{lstlisting}

When \passthrough{\lstinline!OFF!}, the
\passthrough{\lstinline!spectral/!} headers are no-ops
(\passthrough{\lstinline!\#ifdef GRIDCALC\_USE\_FFT!} guards) and the
FD--FFT cross-check fixture is excluded from the test binary. The FD
code path is unaffected.

\subsection{Wavenumber and Nyquist
convention}\label{wavenumber-and-nyquist-convention}

\begin{itemize}
\tightlist
\item
  \textbf{x-axis (rfft):} half-spectrum of length
  \passthrough{\lstinline!Nx/2 + 1!}. Wavenumbers
  \(k_x[n_x] = 2\pi n_x / (N_x h_x)\) for \(n_x \in [0, N_x/2]\).
\item
  \textbf{y, z (full c2c):} signed
  \passthrough{\lstinline!numpy.fft!}-style.
  \(k_a[n_a] = 2\pi n_a^{\text{signed}} / (N_a h_a)\) where
  \(n_a^{\text{signed}} = n_a\) for \(n_a < N_a/2\) else \(n_a - N_a\).
\item
  \textbf{Nyquist zeroing.} For any axis whose per-axis derivative count
  is \textbf{odd}, the mode at \(n_a = N_a/2\) (Nyquist) is set to zero
  before applying the spectral multiplier. This handles the standard
  \(\pm \pi/h\) sign ambiguity at the Nyquist that would otherwise alias
  under the inverse transform.
\end{itemize}

\subsection{Worked example: a CI-style FD--FFT
assertion}\label{worked-example-a-ci-style-fdfft-assertion}

Treat the spectral path as the reference and assert that an FD operator
agrees with it within the operator's truncation order:

\begin{lstlisting}[language={C++}]
#include <cmath>
#include <gridcalc/core/EigenAliases.hpp>
#include <gridcalc/core/Field.hpp>
#include <gridcalc/core/Grid.hpp>
#include <gridcalc/diff/Biharmonic.hpp>
#include <gridcalc/spectral/Derivatives.hpp>

using gridcalc::core::Field;
using gridcalc::core::Grid;
using gridcalc::core::Vec3d;

constexpr double kPi = 3.14159265358979323846;

const int N = 64;
const double L = 2.0 * kPi;
const double h = L / N;

Grid grid(N, N, N, Vec3d(h, h, h));
Field<double> psi(grid, [](double x, double y, double z) {
  return std::sin(x + y + z);
});

Field<double> fd  = gridcalc::diff::biharmonic<4>(psi);
Field<double> sp  = gridcalc::spectral::biharmonic(psi);

// Compute relative max-norm discrepancy:
double max_diff = 0.0;
double max_ref  = 0.0;
for (std::size_t n = 0; n < psi.getSize(); ++n) {
  max_diff = std::max(max_diff, std::abs(fd.data()[n] - sp.data()[n]));
  max_ref  = std::max(max_ref,  std::abs(sp.data()[n]));
}
const double rel = max_diff / max_ref;
// Order-4 truncation at N=64 on |k|=sqrt(3): rel ~ h^4 * |k|^2 ~ 3e-4.
// A loose 1e-3 bound is plenty for a CI gate.
assert(rel < 1e-3);
\end{lstlisting}

The cross-check fixture
\href{../../../test/fd_fft_crosscheck_test.cpp}{\passthrough{\lstinline!test/fd\_fft\_crosscheck\_test.cpp!}}
generalizes this pattern to every Phase 1--8 operator and asserts the
log-log slope on a four-point grid sweep.

\subsection{\texorpdfstring{What this does \emph{not}
do}{What this does not do}}\label{what-this-does-not-do}

\begin{itemize}
\tightlist
\item
  \textbf{Production differentiation.} Use
  \passthrough{\lstinline!gridcalc::diff::*!} instead.
\item
  \textbf{Non-orthogonal Bravais grids.} Phase 15.
\item
  \textbf{Multi-atom-basis grids.} Phase 16.
\item
  \textbf{31 named \passthrough{\lstinline!spectral::*!} d-prefix
  partials.} Use
  \passthrough{\lstinline!spectral::partial<Nx, Ny, Nz>(field)!}
  instead.
\item
  \textbf{Spectral solver path.} Phase 19's implicit-scheme spec round
  will reconsider this; the Phase 9 vendor of PocketFFT is
  verification-only.
\end{itemize}

For the wavenumber sign convention, the Nyquist-aliasing argument, and
the design choices behind the hybrid API and the vendored PocketFFT, see
\href{../../developer-note/notes/phase-9-spectral-verification.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-9-spectral-verification.md!}}.
