\chapter{Periodic 3D scalar Laplacian}
\label{chap:dev:periodic-3d-scalar-laplacian}

\begin{quote}
Phase metadata. Spec dir:
\passthrough{\lstinline!specs/2026-05-01-phase-1-laplacian/!}.
Implements \passthrough{\lstinline!core::Grid!},
\passthrough{\lstinline!core::Field<T, Policy = Periodic>!},
\passthrough{\lstinline!core::IndexPolicy::Periodic!},
\passthrough{\lstinline!stencil::Coefficients<2>!}, and
\passthrough{\lstinline!diff::laplacian!}. Version target:
\passthrough{\lstinline!0.1.0!}.

\emph{This Developer Note was written retroactively after Phase 4, when
the Phase-3-onward doc-notes rule was extended to cover Phases 1 and 2
as well (see specs/STATUS.md ``Decisions worth knowing'').}
\end{quote}

\subsection{1. Theory}\label{theory}

Phase 1 stands up the \textbf{discrete approximation of \(\nabla^2\)} on
a periodic Cartesian domain
\(\Omega = [0, L_x) \times [0, L_y) \times [0, L_z)\) with
\(L_a = N_a h_a\) for \(a \in \{x, y, z\}\). A scalar field
\(\psi : \Omega \to \mathbb{R}\) is sampled at the lattice points
\(\mathbf{x}_{ijk} = (i h_x,\, j h_y,\, k h_z)\) for \(0 \le i < N_x\),
\(0 \le j < N_y\), \(0 \le k < N_z\).

The continuous Laplacian on this domain has tensor-product trig
eigenfunctions: for
\(\psi_{\mathbf{k}}(\mathbf{x}) = \prod_a \sin(k_a x_a)\) with integer
wavenumbers \(k_a \in \mathbb{Z}_{>0}\),

\[
\nabla^2 \psi_{\mathbf{k}} \;=\; -(k_x^2 + k_y^2 + k_z^2)\,\psi_{\mathbf{k}}.
\]

Eigenfunctions of the discrete Laplacian are the \textbf{same trig
products} sampled at the lattice; the discrete eigenvalues differ from
the continuous ones by a quantifiable \(O(h^2)\) correction (derived in
§2). The 3D Laplacian decomposes as a sum of 1D second derivatives along
each axis (Cartesian-orthogonal ⇒ no mixed partials), so the
implementation applies the 1D 2nd-order central-difference stencil along
each axis independently and sums.

\subsection{2. Math derivation}\label{math-derivation}

\subsubsection{1D second-derivative central-difference
stencil}\label{d-second-derivative-central-difference-stencil}

Taylor-expanding \(f(x \pm h)\):

\[
f(x + h) = f(x) + h f'(x) + \tfrac{h^2}{2} f''(x) + \tfrac{h^3}{6} f'''(x) + \tfrac{h^4}{24} f''''(x) + O(h^5),
\]

\[
f(x - h) = f(x) - h f'(x) + \tfrac{h^2}{2} f''(x) - \tfrac{h^3}{6} f'''(x) + \tfrac{h^4}{24} f''''(x) + O(h^5).
\]

Adding cancels the odd derivatives:

\[
f(x + h) + f(x - h) - 2 f(x) \;=\; h^2 f''(x) + \tfrac{h^4}{12} f''''(x) + O(h^6).
\]

Dividing by \(h^2\) gives the \textbf{2nd-order accurate stencil} that
\passthrough{\lstinline!stencil::Coefficients<2>!} encodes:

\[
\partial^2_x f(x) \;\approx\; \frac{f(x - h) - 2 f(x) + f(x + h)}{h^2}
\;=\; f''(x) + \tfrac{h^2}{12} f''''(x) + O(h^4),
\]

with \textbf{truncation error \(-\tfrac{h^2}{12} f''''(x)\)} (using the
discrete-minus-continuous sign convention). Weights
\passthrough{\lstinline!\{1, -2, 1\}!} divided by \(h^2\), encoded as
the 3-element table
\passthrough{\lstinline!Coefficients<2>::\{offsets, weights\}!} per
\href{../../../include/gridcalc/stencil/CentralDifference.hpp}{\passthrough{\lstinline!include/gridcalc/stencil/CentralDifference.hpp!}}.
The \passthrough{\lstinline!Order = 4!} specialization is deferred to
Phase 7.

\subsubsection{3D Laplacian via Cartesian
decomposition}\label{d-laplacian-via-cartesian-decomposition}

The 3D operator is a sum of three 1D operators, one per axis, each
scaled by its own \(1/h_a^2\):

\[
\nabla^2_h \psi_{ijk} \;=\; \sum_{a \in \{x, y, z\}} \frac{1}{h_a^2} \sum_{m=-1}^{+1} w_m\, \psi(\mathbf{x}_{ijk} + m h_a \hat{\mathbf{e}}_a),
\]

with \(w = (1, -2, 1)\). This is implemented as three nested loops in
\href{../../../include/gridcalc/diff/Laplacian.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Laplacian.hpp!}}
that each accumulate into a per-point scalar before writing to the
output \passthrough{\lstinline!Field!}.

\subsubsection{Discrete eigenvalue on a trig
product}\label{discrete-eigenvalue-on-a-trig-product}

For \(\psi_{ijk} = \sin(k_x x_i) \sin(k_y y_j) \sin(k_z z_k)\) on the
periodic grid (with each \(k_a\) a positive integer so that the function
is grid-periodic):

\[
\frac{\sin(k_a (x_a - h_a)) - 2\sin(k_a x_a) + \sin(k_a (x_a + h_a))}{h_a^2} \;=\; -\frac{2(1 - \cos(k_a h_a))}{h_a^2}\,\sin(k_a x_a),
\]

so \(\psi_{\mathbf{k}}\) is an exact eigenfunction of \(\nabla^2_h\)
with eigenvalue

\[
\lambda_h(\mathbf{k}) \;=\; -\sum_a \frac{2(1 - \cos(k_a h_a))}{h_a^2}.
\]

Taylor-expanding
\(1 - \cos(k_a h_a) = (k_a h_a)^2/2 - (k_a h_a)^4/24 + O(h^6)\) gives

\[
\lambda_h(\mathbf{k}) \;=\; -\sum_a k_a^2 \;+\; \frac{1}{12} \sum_a k_a^4 h_a^2 \;+\; O(h^4),
\]

i.e.~the \textbf{continuous eigenvalue \(-\sum_a k_a^2\) plus an
\(O(h^2)\) correction} with leading coefficient
\(\tfrac{1}{12} \sum_a k_a^4\). For the test
\passthrough{\lstinline!LaplacianTest.RecoversTrigEigenvalue!} at
\(N = 32\), \(h = 2\pi/32\), \(\mathbf{k} = (1, 2, 1)\):

\[
\frac{|\lambda_h - \lambda|}{|\lambda|} \;\approx\; \frac{(1^4 + 2^4 + 1^4)\,h^2}{12 \cdot (1^2 + 2^2 + 1^2)} \;=\; \frac{18 h^2}{72} \;=\; \frac{h^2}{4} \;\approx\; 9.6 \times 10^{-3}.
\]

This matches the test's tolerance of \passthrough{\lstinline!1.5e-2!}
with healthy headroom and is the per-axis-independent error model the
convergence-sweep test exploits at \(N \in \{16, 32, 64\}\).

\subsection{3. Algorithm}\label{algorithm}

\subsubsection{Grid}\label{grid}

\href{../../../include/gridcalc/core/Grid.hpp}{\passthrough{\lstinline!core::Grid!}}
is a small POD holding \((N_x, N_y, N_z)\) and
\passthrough{\lstinline!Vec3d cell\_size!} \((h_x, h_y, h_z)\).
\passthrough{\lstinline!getLinearIndex(i, j, k)!} returns the i-fastest
offset \passthrough{\lstinline!i + Nx*(j + Ny*k)!}. Construction
validates that every extent and every cell-size component is strictly
positive.

\subsubsection{Field\textless T,
Policy\textgreater{}}\label{fieldt-policy}

\href{../../../include/gridcalc/core/Field.hpp}{\passthrough{\lstinline!core::Field<T, Policy>!}}
holds a \passthrough{\lstinline!Grid!} by value and a
\passthrough{\lstinline!std::vector<T>!} of length \(N_x N_y N_z\) in
i-fastest layout. Three constructors:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \passthrough{\lstinline!Field(grid)!} --- zero-initialized via
  \passthrough{\lstinline!std::vector<T>(N)!} value-init.
\item
  \passthrough{\lstinline!Field(grid, value)!} --- broadcast.
\item
  \passthrough{\lstinline!Field(grid, f)!} --- SFINAE-enabled when
  \passthrough{\lstinline!f!} is invocable as
  \passthrough{\lstinline!f(double, double, double) -> T!}; samples
  \passthrough{\lstinline!f!} at the Cartesian coordinates
  \passthrough{\lstinline!(i*hx, j*hy, k*hz)!}.
\end{enumerate}

\passthrough{\lstinline!operator()(i, j, k)!} always passes its
arguments through \passthrough{\lstinline!Policy::wrap!} before calling
\passthrough{\lstinline!Grid::getLinearIndex!}. For the default
\passthrough{\lstinline!IndexPolicy::Periodic!}, this is the modular
wrap \passthrough{\lstinline!((i \% N) + N) \% N!}. The wrap is always
applied, never special-cased on whether \passthrough{\lstinline!i!} is
in range --- branchless modular arithmetic in the hot path is the right
tradeoff because it eliminates the branch predictor miss that a range
check would incur on every stencil access.

\subsubsection{Laplacian operator}\label{laplacian-operator}

\href{../../../include/gridcalc/diff/Laplacian.hpp}{\passthrough{\lstinline!diff::laplacian!}}
loops over all grid points (z-slowest, x-fastest, matching the storage
layout). At each point, it accumulates three independent
\passthrough{\lstinline!Coefficients<2>!}-weighted sums --- one per axis
--- and returns
\passthrough{\lstinline!dxx*inv\_hx2 + dyy*inv\_hy2 + dzz*inv\_hz2!}.
Because \passthrough{\lstinline!operator()!} already wraps via the
periodic policy, no separate boundary handling code path is needed; the
same loop body works for boundary and interior points.

Algorithmic complexity is \(O(N_x N_y N_z)\). The output is a fresh
\passthrough{\lstinline!Field<double>!}, allocated once at the start.

\subsection{4. Design decisions}\label{design-decisions}

\begin{itemize}
\tightlist
\item
  \textbf{\passthrough{\lstinline!Field<T, Policy>!} is templated on the
  policy from Phase 1.} Only \passthrough{\lstinline!Periodic!} exists
  at Phase 1; \passthrough{\lstinline!Dirichlet!} and
  \passthrough{\lstinline!Neumann!} arrive in Phase 18. The policy
  template parameter was the \emph{only} piece of forward-looking design
  admitted at Phase 1 --- specifically so Phase 18 is an additive
  change, not a public-API break. See
  \href{../../../specs/2026-05-01-phase-1-laplacian/requirements.md}{Phase
  1 requirements.md}.
\item
  \textbf{i-fastest storage layout.} Linear offset
  \passthrough{\lstinline!i + Nx*(j + Ny*k)!}. Matches PocketFFT / FFTW
  (Phase 9) and Eigen's column-major default. Public-contract status
  from Phase 1 forward; Phase 20 benchmarks would be invalidated by a
  layout flip later.
\item
  \textbf{\passthrough{\lstinline!stencil::Coefficients<Order>!} is a
  template; primary undefined.} Only \passthrough{\lstinline!Order = 2!}
  is specialized at Phase 1; instantiating an unsupported
  \passthrough{\lstinline!Order!} trips a compile error at the call site
  rather than silently producing zero output. Phase 7 adds the
  \passthrough{\lstinline!Order = 4!} specialization with no call-site
  refactor.
\item
  \textbf{Eigen propagated as a \passthrough{\lstinline!SYSTEM!}
  include.} Required so \passthrough{\lstinline!-Wconversion!} (and the
  rest of the strict warning set) stays active on the project's own code
  without firing on Eigen's NEON typecasting headers.
  \passthrough{\lstinline!WARNINGS\_AS\_ERRORS=ON!} in CI breaks without
  this. See the \passthrough{\lstinline!tech-stack.md!} Eigen section.
\item
  \textbf{\passthrough{\lstinline!Vec3d!} originally lived in
  \passthrough{\lstinline!core/Grid.hpp!}.} Phase 2 relocated the alias
  to \passthrough{\lstinline!core/EigenAliases.hpp!} once
  \passthrough{\lstinline!gradient!} and
  \passthrough{\lstinline!divergence!} also needed it; the
  fully-qualified name \passthrough{\lstinline!gridcalc::core::Vec3d!}
  is unchanged.
\end{itemize}

\subsection{5. References}\label{references}

{[}1{]} LeVeque, R. J. (2007). \emph{Finite Difference Methods for
Ordinary and Partial Differential Equations}. SIAM. ISBN
978-0-89871-783-9. \url{https://doi.org/10.1137/1.9780898717839}. §1.2
on second-derivative central differences and the truncation analysis
used in §2 of this note; §2 on Cartesian-product extension of 1D
stencils to higher dimensions.

{[}2{]} Trefethen, L. N. (2000). \emph{Spectral Methods in MATLAB}.
SIAM. ISBN 978-0-89871-465-4.
\url{https://doi.org/10.1137/1.9780898719598}. Chapter 1 covers the
spectral relationship between trig basis functions and finite
differences on periodic domains, the framing used in §1 of this note.

{[}3{]} cppreference: \passthrough{\lstinline!std::vector!}
value-initialization semantics.
\url{https://en.cppreference.com/w/cpp/container/vector/vector}
(permanent URL). Confirms the zero-initialization guarantee for
\passthrough{\lstinline!Field(grid)!} when \passthrough{\lstinline!T!}
is \passthrough{\lstinline!double!}. Note: this guarantee does
\emph{not} hold for Eigen fixed-size types (the default constructor
leaves coefficients uninitialized) --- Phase 2's tests work around this
by explicitly broadcast-initializing
\passthrough{\lstinline!Field<Vec3d>!} with
\passthrough{\lstinline!Vec3d::Zero()!}.
