\chapter{Functional evaluation}
\label{chap:dev:functional-evaluation}

\begin{quote}
Phase metadata. Spec dir:
\passthrough{\lstinline!specs/2026-05-02-phase-4-functional-evaluation/!}.
Implements \passthrough{\lstinline!gridcalc::func::evaluate!}. Version
target: \passthrough{\lstinline!0.4.0!}.
\end{quote}

\subsection{1. Theory}\label{theory}

A \textbf{functional} maps a function (here, a scalar field) to a real
number:

\[
F[\psi] \;=\; \int_{\Omega} f\bigl(\psi(\mathbf{x}),\, \nabla\psi(\mathbf{x}),\, \nabla^2\psi(\mathbf{x})\bigr)\, dV,
\]

with \(\psi : \Omega \to \mathbb{R}\) on the periodic Cartesian domain
\(\Omega = [0, L_x) \times [0, L_y) \times [0, L_z)\) and \(f\) a smooth
density. Functionals of this form arise throughout continuum and
statistical physics: Ginzburg--Landau and Cahn--Hilliard free energies,
the Helfrich bending energy of membranes, kinetic-energy functionals in
density-functional theory, and many phase-field models. Phase 4 stands
up the \textbf{scalar-field, periodic} version; vector / tensor /
non-periodic extensions arrive in later phases.

The \textbf{discrete functional} \(F_h\) replaces the integral with the
periodic midpoint quadrature from Phase 3 and the continuous derivatives
with the 2nd-order central differences from Phases 1 and 2:

\[
F_h(\psi) \;=\; \Delta V \sum_{i,j,k} f\bigl(\psi_{ijk},\, (\nabla_h \psi)_{ijk},\, (\nabla^2_h \psi)_{ijk}\bigr).
\]

The convergence of \(F_h \to F\) as \(h \to 0\) is bounded by the
\textbf{slowest component}. The integration is spectrally accurate for
analytic periodic integrands (Phase 3, reference {[}3{]}). The gradient
and Laplacian stencils are 2nd-order. The composed convergence rate is
therefore \(O(h^2)\) \textbf{whenever the integrand actually consumes a
derivative}; integrands that depend on \(\psi\) alone inherit the
spectrally accurate quadrature error directly.

\subsection{2. Math derivation}\label{math-derivation}

\subsubsection{Convergence}\label{convergence}

For a smooth functional density \(f\), Taylor-expand around the exact
derivatives:

\[
f(\psi, \nabla_h\psi, \nabla^2_h\psi) \;=\; f(\psi, \nabla\psi, \nabla^2\psi) \;+\; \partial_{(\nabla\psi)} f \cdot (\nabla_h - \nabla)\psi \;+\; \partial_{(\nabla^2\psi)} f \cdot (\nabla^2_h - \nabla^2)\psi \;+\; \text{higher order}.
\]

The pointwise gradient error is \((\nabla_h - \nabla)\psi = O(h^2)\)
(Phase 2 stencil); the pointwise Laplacian error is \(O(h^2)\) (Phase 1
stencil); and integrating an \(O(h^2)\) pointwise error over a finite
domain produces an \(O(h^2)\) functional error (the integration is
spectrally accurate for the smooth periodic prefactor). For the
Ginzburg--Landau example below, both the leading-order constant and the
\(O(h^2)\) convergence are matched by experiment to within 5\%.

\subsubsection{Hand-computed reference for
Ginzburg--Landau}\label{hand-computed-reference-for-ginzburglandau}

For \(\psi = \sin(x)\sin(y)\sin(z)\) on \([0, 2\pi]^3\) and \(W = 1\):

\textbf{Gradient term} \(\int (\tfrac{1}{2})|\nabla\psi|^2\, dV\):

\[
|\nabla\psi|^2 \;=\; \cos^2 x \sin^2 y \sin^2 z \;+\; \sin^2 x \cos^2 y \sin^2 z \;+\; \sin^2 x \sin^2 y \cos^2 z.
\]

Using \(\int_0^{2\pi}\sin^2 = \int_0^{2\pi}\cos^2 = \pi\), each
axis-aligned \(\cos^2\sin^2\sin^2\) tensor product integrates to
\(\pi \cdot \pi \cdot \pi = \pi^3\). Three terms total give \(3\pi^3\),
hence

\[
\int \tfrac{1}{2}|\nabla\psi|^2\, dV \;=\; \tfrac{3}{2}\pi^3.
\]

\textbf{Potential term}
\(\int (\psi^2 - 1)^2\, dV = \int [\psi^4 - 2\psi^2 + 1]\, dV\):

\begin{itemize}
\tightlist
\item
  \(\int \psi^4\, dV\) uses \(\int_0^{2\pi}\sin^4 = 3\pi/4\), so
  \(\int \sin^4 x \sin^4 y \sin^4 z\, dV = (3\pi/4)^3 = 27\pi^3/64\).
\item
  \(\int \psi^2\, dV = \pi \cdot \pi \cdot \pi = \pi^3\).
\item
  \(\int 1\, dV = (2\pi)^3 = 8\pi^3\).
\item
  Sum:
  \(27\pi^3/64 - 2\pi^3 + 8\pi^3 = 27\pi^3/64 + 6\pi^3 = 411\pi^3/64\).
\end{itemize}

\textbf{Total} (with \(W = 1\)):

\[
F = \tfrac{3}{2}\pi^3 \;+\; \tfrac{411}{64}\pi^3 \;=\; \tfrac{507}{64}\pi^3 \;\approx\; 245.62.
\]

\subsubsection{Why the 1-arg path is
exact}\label{why-the-1-arg-path-is-exact}

When \(f\) depends only on \(\psi\), no stencil is applied ---
\(f(\psi_{ijk})\) is read directly from the input field, and the only
error comes from the Phase 3 periodic-trapezoidal rule. For analytic
periodic integrands the quadrature converges exponentially in \(h\)
(reference {[}3{]}, Theorem 3.2), so the test
\href{../../../test/functional_test.cpp}{\passthrough{\lstinline!FunctionalTest.PsiOnlyArity!}}
sets its tolerance to \(10^{-12}\) rather than the \(10^{-2}\) used by
the gradient-bearing tests.

\subsection{3. Algorithm}\label{algorithm}

The implementation lives in
\href{../../../include/gridcalc/func/Functional.hpp}{\passthrough{\lstinline!include/gridcalc/func/Functional.hpp!}}.

\begin{lstlisting}[language={C++}]
template <class F, class Tag = Pairwise>
inline double evaluate(const core::Field<double>& psi, F&& f, Tag tag = {});
\end{lstlisting}

\subsubsection{Arity dispatch}\label{arity-dispatch}

The key piece is an \passthrough{\lstinline!if constexpr!} ladder on
\passthrough{\lstinline!std::is\_invocable\_r\_v<double, F\&, …>!}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \passthrough{\lstinline!f(double, const Vec3d\&, double)!} matches →
  materialize \textbf{both}
  \passthrough{\lstinline!diff::gradient(psi)!} and
  \passthrough{\lstinline!diff::laplacian(psi)!}.
\item
  else \passthrough{\lstinline!f(double, const Vec3d\&)!} matches →
  materialize \textbf{only}
  \passthrough{\lstinline!diff::gradient(psi)!}.
\item
  else \passthrough{\lstinline!f(double)!} matches → no stencil
  materialization.
\item
  else trip a \passthrough{\lstinline!static\_assert!} via the
  \passthrough{\lstinline!detail::deferred\_false<F>!} trick.
\end{enumerate}

Listing the longest-arity branch first means a callable with default
arguments (which would match all three signatures) selects the full
triplet, which is the safer interpretation --- the user provided
defaults specifically to allow the extra arguments.

The \passthrough{\lstinline!detail::deferred\_false<F>!} indirection is
required because plain
\passthrough{\lstinline!static\_assert(false, …)!} inside an
\passthrough{\lstinline!if constexpr!} else-branch fires unconditionally
under C++17 (resolved in C++23 by CWG2518; reference {[}4{]}).

\subsubsection{Eager materialization}\label{eager-materialization}

Each arity branch:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Calls \passthrough{\lstinline!diff::gradient(psi)!} and/or
  \passthrough{\lstinline!diff::laplacian(psi)!} as needed. These return
  fresh \passthrough{\lstinline!Field<Vec3d>!} and
  \passthrough{\lstinline!Field<double>!} respectively, so total
  temporary storage is up to \(4 N_x N_y N_z\) doubles (3 for the
  gradient field's Vec3d entries, 1 for the Laplacian).
\item
  Allocates a \passthrough{\lstinline!Field<double> integrand(grid)!} of
  the same size.
\item
  Iterates \passthrough{\lstinline!(k, j, i)!} (z-slowest, x-fastest ---
  matches the i-fastest storage layout from Phase 1) and writes
  \passthrough{\lstinline!f(...)!} into
  \passthrough{\lstinline!integrand!}.
\item
  Returns \passthrough{\lstinline!func::integrate(integrand, tag)!}.
\end{enumerate}

Algorithmic complexity is \(O(N_x N_y N_z)\). Memory is
\(O(N_x N_y N_z)\) on top of the input.

\subsubsection{Tag forwarding}\label{tag-forwarding}

\passthrough{\lstinline!evaluate!} takes a \passthrough{\lstinline!Tag!}
defaulted to \passthrough{\lstinline!Pairwise\{\}!} and forwards it to
\passthrough{\lstinline!integrate!}. A
\passthrough{\lstinline!static\_assert!} at the top of the function
constrains \passthrough{\lstinline!Tag!} to
\passthrough{\lstinline!Pairwise!} or \passthrough{\lstinline!Kahan!} so
a misspelled tag fails at the call site with a readable diagnostic
instead of an unrelated overload-set error from
\passthrough{\lstinline!integrate!}.

\subsubsection{Tests}\label{tests}

\passthrough{\lstinline!test/functional\_test.cpp!} exercises all three
dispatch paths plus the Ginzburg--Landau hand-computed reference and a
2nd-order convergence sweep.

\subsection{4. Design decisions}\label{design-decisions}

\begin{itemize}
\tightlist
\item
  \textbf{SFINAE-detected arity rather than the strict roadmap
  signature.} The roadmap specifies
  \passthrough{\lstinline!f(double, Vector3d, double)!} only; Phase 4
  widens this to a strict superset (1-, 2-, or 3-arg) because the
  dispatch is trivial in \passthrough{\lstinline!if constexpr!}, the
  user gets ergonomically natural signatures, and unused stencils are
  skipped --- a real correctness and performance win even at Phase 4.
  See the
  \href{../../../specs/2026-05-02-phase-4-functional-evaluation/requirements.md}{\passthrough{\lstinline!requirements.md!}}
  ``Decisions made for this feature'' entry.
\item
  \textbf{Eager materialization over per-point lazy evaluation.}
  Per-point lazy stencils would save the gradient and Laplacian field
  allocations but duplicate the stencil formulas inline, blocking direct
  reuse of \passthrough{\lstinline!diff::gradient!} /
  \passthrough{\lstinline!diff::laplacian!}. Phase 11+ has a stronger
  reason to revisit this (rank-extended functionals where the rank-6
  Hessian of a rank-2 field would explode memory if materialized).
\item
  \textbf{Vec3d argument by \passthrough{\lstinline!const!} reference.}
  Per-point Vec3d allocation is cheap on the stack but copying the 24
  bytes is wasteful in the inner loop. The SFINAE check uses
  \passthrough{\lstinline!const Vec3d\&!} so user lambdas can declare it
  either way (by-value lambdas accept the const-ref argument and copy).
\item
  \textbf{Worked example is Ginzburg--Landau} rather than a simpler
  integrand: matches the roadmap deliverable and exercises the 2-arg
  dispatch path, which is the most common arity in practice
  (gradient-energy functionals).
\end{itemize}

\subsection{5. References}\label{references}

{[}1{]} Provatas, N., \& Elder, K. (2010). \emph{Phase-Field Methods in
Materials Science and Engineering}. Wiley-VCH. ISBN 978-3-527-40747-7.
Chapter 3 introduces the Ginzburg--Landau functional in the form
\(F = \int (\tfrac{1}{2})|\nabla\psi|^2 + W\,(\psi^2-1)^2\, dV\) used as
the worked example here. \url{https://doi.org/10.1002/9783527631520}.

{[}2{]} Boyd, J. P. (2001). \emph{Chebyshev and Fourier Spectral
Methods} (2nd ed.). Dover. ISBN 978-0-486-41183-5. Chapter 4 covers
tensor-product quadrature and the inheritance of convergence rates from
the slowest-component approximation, the argument used in §1 of this
note. Free PDF at the author's site:
\url{https://websites.umich.edu/~jpboyd/BOOK_Spectral2000.html}
(permanent URL).

{[}3{]} Trefethen, L. N., \& Weideman, J. A. C. (2014). ``The
Exponentially Convergent Trapezoidal Rule.'' \emph{SIAM Review}, 56(3),
385--458. \url{https://doi.org/10.1137/130932132}. Used here for the
spectral accuracy of the integration step (cited in Phase 3's Developer
Note as well).

{[}4{]} cppreference: \passthrough{\lstinline!std::is\_invocable\_r!}.
\url{https://en.cppreference.com/w/cpp/types/is_invocable} (permanent
URL). Provides the precise semantics of the SFINAE check used in the
arity dispatch.

{[}5{]} CWG Issue 2518 (C++ Core Working Group). ``Conformance
requirements and \passthrough{\lstinline!static\_assert(false)!} in
\passthrough{\lstinline!if constexpr!}.''
\url{https://www.open-std.org/jtc1/sc22/wg21/docs/cwg_defects.html\#2518}
(permanent URL). Documents the C++23 resolution that allows
\passthrough{\lstinline!static\_assert(false, …)!} inside
\passthrough{\lstinline!if constexpr!} without the
\passthrough{\lstinline!deferred\_false!} workaround. Phase 4 still
targets C++17 per \passthrough{\lstinline!tech-stack.md!}, so the
workaround is required.
