\chapter{Higher-order spatial derivatives}
\label{chap:dev:higher-order-spatial-derivatives}

\begin{quote}
Phase metadata. Spec dir:
\passthrough{\lstinline!specs/2026-05-02-phase-8-higher-order-derivatives/!}.
Adds \passthrough{\lstinline!stencil::ThirdDerivative<Order>!},
\passthrough{\lstinline!stencil::FourthDerivative<Order>!} (orders 2 and
4 each), the \passthrough{\lstinline!diff::detail::mixedDerivative!}
tensor-product helper, 31 named partial-derivative functions across
\passthrough{\lstinline!diff/MixedPartial.hpp!} /
\passthrough{\lstinline!diff/ThirdOrder.hpp!} /
\passthrough{\lstinline!diff/FourthOrder.hpp!}, and
\passthrough{\lstinline!diff::biharmonic!} with the
\passthrough{\lstinline!diff::d4!} alias. Version target:
\passthrough{\lstinline!0.8.0!}.
\end{quote}

\subsection{1. Theory}\label{theory}

\subsubsection{Multi-index partials on a Cartesian-orthogonal periodic
grid}\label{multi-index-partials-on-a-cartesian-orthogonal-periodic-grid}

Let \(\boldsymbol\alpha = (n_x, n_y, n_z) \in \mathbb{Z}_{\geq 0}^3\) be
a multi-index with total order
\(|\boldsymbol\alpha| = n_x + n_y + n_z\). The partial derivative

\[
\partial^{\boldsymbol\alpha}\psi \;=\; \frac{\partial^{|\boldsymbol\alpha|}\psi}{\partial x^{n_x}\,\partial y^{n_y}\,\partial z^{n_z}}
\]

is well-defined for \(\psi \in C^{|\boldsymbol\alpha|}\). On the
\textbf{Cartesian- orthogonal} grid that Phase 1 ships
(\(\Omega = [0,L_x) \times [0,L_y) \times
[0,L_z)\) with periodic policy and per-axis cell sizes
\(h_x, h_y, h_z\)), the derivative operator is \textbf{separable}:

\[
\partial^{\boldsymbol\alpha} \;=\; \partial_x^{n_x}\circ\partial_y^{n_y}\circ\partial_z^{n_z},
\]

i.e., the multi-axis derivative is the composition (in any order) of the
per-axis 1D derivatives. Discretely this means the \textbf{stencil for
\(\partial^{\boldsymbol\alpha}\) on the grid factors as a tensor
product} of three 1D stencils:

\[
(D^{\boldsymbol\alpha}\psi)_{ijk} \;=\; \frac{1}{h_x^{n_x} h_y^{n_y} h_z^{n_z}}\,\sum_{a, b, c} w^{(n_x)}_a\,w^{(n_y)}_b\,w^{(n_z)}_c\,\psi_{i+a,\,j+b,\,k+c},
\]

where \(\{w^{(n)}_m\}\) is the 1D central-difference weight table for
\(\partial^n / \partial x^n\) at the chosen accuracy order. \textbf{The
truncation order of the multi-axis stencil equals the minimum of the
per-axis truncation orders} --- proof sketch: the leading error of the
per-axis 1D stencil is \(O(h^p)\) with \(p\) the per-axis order;
tensoring gives error that is the sum of three such per-axis terms, each
\(O(h_a^{p})\), hence the total order is \(\min_a p_a\). Since this
phase ships every per-axis table at the \textbf{same} target order
(\passthrough{\lstinline!Order!} parameter), the multi-axis stencil
inherits exactly that order.

\subsubsection{The biharmonic as a contracted scalar
operator}\label{the-biharmonic-as-a-contracted-scalar-operator}

The biharmonic is defined by \textbf{contraction} rather than by the
tensor-product form above:

\[
\nabla^4 \;=\; (\nabla^2)^2 \;=\; \biggl(\sum_a \partial_a^2\biggr)\biggl(\sum_b \partial_b^2\biggr) \;=\; \sum_a \partial_a^4 \;+\; 2\sum_{a < b} \partial_a^2\,\partial_b^2.
\]

In 3D this is \textbf{3 single-axis 4th derivatives + 3 mixed (2,2)
cross-terms} (the cross-terms picking up a factor of 2 from
\(\partial_a^2\partial_b^2 +
\partial_b^2\partial_a^2\) in the expansion). Phase 8's
\passthrough{\lstinline!biharmonic!} implements exactly this six-term
sum directly (not via \passthrough{\lstinline!laplacian∘laplacian!});
the ``Algorithm'' section explains why.

For \(\psi(\mathbf{x}) = \sin(\mathbf{k}\cdot\mathbf{x})\), every
\(\partial_a^2\psi = -k_a^2\psi\), so
\(\nabla^2\psi = -|\mathbf{k}|^2\psi\) and
\(\nabla^4\psi = |\mathbf{k}|^4\psi\). This is the roadmap acceptance
bar.

\subsubsection{Symmetry constraints on the new 1D
tables}\label{symmetry-constraints-on-the-new-1d-tables}

For an \textbf{odd} total derivative (ranks 1, 3, 5, \ldots), the weight
table is \textbf{anti-symmetric}: \(w_{-m} = -w_m\), so \(w_0 = 0\). For
an \textbf{even} total derivative (ranks 2, 4, 6, \ldots), the table is
\textbf{symmetric}: \(w_{-m} = +w_m\), \(w_0\) free. Doubling the
half-width \(r\) buys two more matched Taylor orders by symmetry ---
odd-order error terms cancel automatically. So central differences come
in \textbf{even} truncation orders only (2, 4, 6, \ldots) and there is
no order-3 symmetric stencil for these derivatives.

\subsection{2. Math derivation}\label{math-derivation}

\subsubsection{Order-2 third-derivative stencil (5-point,
anti-symmetric)}\label{order-2-third-derivative-stencil-5-point-anti-symmetric}

We seek \(\partial^3 f / \partial x^3\) at \(x_i\) from

\[
\partial_x^3 f(x_i) \;\approx\; \frac{1}{h^3}\sum_{m=-2}^{2} w_m\,f(x_{i+m}),
\]

with anti-symmetry \(w_{-m} = -w_m\) giving \(w_0 = 0\) and two free
parameters \(w_1, w_2\). Taylor-expanding \(f(x_i + m h)\) and grouping
by derivative order, the moment

\[
M_p \;=\; \sum_m w_m\,m^p
\]

vanishes for \textbf{even} \(p\) by anti-symmetry, leaving only
odd-\(p\) moment conditions. For an order-2 approximation of
\(\partial^3 f/\partial x^3\):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3784}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.5135}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
moment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
condition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
reason
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(M_1 = 2 w_1 + 4 w_2\) & \(= 0\) & kill \(f'(x_i)\) leak \\
\(M_3 = 2 w_1 + 16 w_2\) & \(= 6 = 3!\) & match \(h^3 \cdot f'''/3!\) \\
\end{longtable}
}

Solving: \(w_1 - w_2 \cdot 6 = -3\) from \(M_3 - 3 M_1\), simultaneous
with \(M_1 = 0\) gives \(w_2 = 1/2,\ w_1 = -1\). Anti-symmetry then
fills in \(w_{-1} = +1,\ w_{-2} = -1/2\). The leading uncancelled moment
is \(M_5\), which sets the truncation error at \(O(h^2)\).

\[
\boxed{\,\mathtt{ThirdDerivative}\langle 2\rangle\mathtt{::weights} \;=\; \{-1/2,\ +1,\ 0,\ -1,\ +1/2\}.\,}
\]

\subsubsection{Order-4 third-derivative stencil (7-point,
anti-symmetric)}\label{order-4-third-derivative-stencil-7-point-anti-symmetric}

The half-width grows to \(r = 3\); three free parameters
\(w_1, w_2, w_3\). Order-4 accuracy demands \(M_1 = 0\), \(M_3 = 6\),
\(M_5 = 0\) (the next uncancelled odd moment is suppressed). Solving the
resulting \(3 \times 3\) linear system yields

\[
\boxed{\,\mathtt{ThirdDerivative}\langle 4\rangle\mathtt{::weights} \;=\; \{1/8,\ -1,\ 13/8,\ 0,\ -13/8,\ +1,\ -1/8\}.\,}
\]

\subsubsection{Order-2 fourth-derivative stencil (5-point,
symmetric)}\label{order-2-fourth-derivative-stencil-5-point-symmetric}

For \(\partial^4 f/\partial x^4\) with symmetry \(w_{-m} = w_m\), the
moment conditions for an order-2 approximation are \(M_0 = 0\),
\(M_2 = 0\), \(M_4 = 24 = 4!\). Solving gives

\[
\boxed{\,\mathtt{FourthDerivative}\langle 2\rangle\mathtt{::weights} \;=\; \{+1,\ -4,\ +6,\ -4,\ +1\}.\,}
\]

(This is the well-known 5-point fourth-difference stencil; same family
that appears as the 1D part of the 13-point biharmonic stencil and as
the second iterate of the 3-point Laplacian stencil.)

\subsubsection{Order-4 fourth-derivative stencil (7-point,
symmetric)}\label{order-4-fourth-derivative-stencil-7-point-symmetric}

Half-width \(r = 3\); free \(w_0, w_1, w_2, w_3\). Conditions:
\(M_0 = 0\), \(M_2 = 0\), \(M_4 = 24\), \(M_6 = 0\). Solving:

\[
\boxed{\,\mathtt{FourthDerivative}\langle 4\rangle\mathtt{::weights} \;=\; \{-1/6,\ +2,\ -13/2,\ +28/3,\ -13/2,\ +2,\ -1/6\}.\,}
\]

The truncation error is set by the next uncancelled even moment \(M_8\),
at order \(O(h^4)\).

\subsubsection{Closed-form analytical reference for the convergence
sweep}\label{closed-form-analytical-reference-for-the-convergence-sweep}

For \(\psi(\mathbf{x}) = \sin x \sin y \sin z\) on the periodic
\([0, 2\pi)^3\) box, the partial derivative

\[
\partial^{\boldsymbol\alpha}\psi(x, y, z) \;=\; f_{n_x}(x)\,f_{n_y}(y)\,f_{n_z}(z),
\]

where \(f_n\) is the \(n\)-th derivative of \(\sin\):

\[
f_n(\xi) \;=\; \begin{cases}\sin\xi & n \equiv 0\pmod 4,\\ \cos\xi & n \equiv 1\pmod 4,\\ -\sin\xi & n \equiv 2\pmod 4,\\ -\cos\xi & n \equiv 3\pmod 4.\end{cases}
\]

This closed form drives the analytical reference in
\href{../../../test/phase_8_test.cpp}{\passthrough{\lstinline!test/phase\_8\_test.cpp!}},
eliminating the need to call any of the operators under test as their
own reference. For each (operator, Order) pair, the relative max-norm
error against this reference is computed at \(N \in \{16, 32, 64\}\) and
the log-log slope vs \(h\) must lie in
\([\text{Order} - 0.5, \text{Order} + 0.5]\).

\subsection{3. Algorithm}\label{algorithm}

\subsubsection{\texorpdfstring{\texttt{detail::mixedDerivative\textless{}Nx,\ Ny,\ Nz,\ Order\textgreater{}}
(the
helper)}{detail::mixedDerivative\textless Nx, Ny, Nz, Order\textgreater{} (the helper)}}\label{detailmixedderivativenx-ny-nz-order-the-helper}

Defined in
\href{../../../include/gridcalc/diff/detail/MultiIndexDerivative.hpp}{\passthrough{\lstinline!include/gridcalc/diff/detail/MultiIndexDerivative.hpp!}}.
Implementation outline:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The internal trait \passthrough{\lstinline!AxisStencil<N, Order>!}
  maps the per-axis derivative count \passthrough{\lstinline!N!} to the
  corresponding 1D \passthrough{\lstinline!gridcalc::stencil::*!} table
  (\passthrough{\lstinline!N = 0!} is identity;
  \passthrough{\lstinline!N ∈ \{1,2,3,4\}!} aliases
  \passthrough{\lstinline!FirstDerivative!},
  \passthrough{\lstinline!Coefficients!},
  \passthrough{\lstinline!ThirdDerivative!},
  \passthrough{\lstinline!FourthDerivative!} via inheritance).
\item
  The per-grid-point work is a triple-nested loop over the outer-product
  stencil offsets, accumulating a single double
  \passthrough{\lstinline!sum!}. Each loop iteration reads one neighbor
  via \passthrough{\lstinline!psi(i + ox, j + oy, k + oz)!}, which
  delegates periodic wrap to \passthrough{\lstinline!Field::Policy!}.
  Per-axis scaling \(1/h_a^{N_a}\) is computed once outside the
  per-point loop.
\item
  Per-point work is \(\prod_a (2 r_a + 1)\) reads, where \(r_a\) is the
  radius of axis-\(a\)'s table (radius \passthrough{\lstinline!0!} for
  \(N_a = 0\), radii from the chosen \passthrough{\lstinline!Order!}'s
  tables otherwise).
\end{enumerate}

The named functions in \passthrough{\lstinline!MixedPartial.hpp!},
\passthrough{\lstinline!ThirdOrder.hpp!}, and
\passthrough{\lstinline!FourthOrder.hpp!} are 2-line wrappers around
this helper, instantiated with the right multi-index. There is exactly
one site of correctness for the multi-axis stencil math.

\subsubsection{\texorpdfstring{\texttt{biharmonic\textless{}Order\textgreater{}(psi)}
(the contracted
operator)}{biharmonic\textless Order\textgreater(psi) (the contracted operator)}}\label{biharmonicorderpsi-the-contracted-operator}

Defined in
\href{../../../include/gridcalc/diff/Biharmonic.hpp}{\passthrough{\lstinline!include/gridcalc/diff/Biharmonic.hpp!}}.
Implementation:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Single fused pass over the grid} --- not
  \passthrough{\lstinline!laplacian(laplacian(psi))!}. The composition
  through an intermediate \passthrough{\lstinline!Field!} would inherit
  the stride-2 leading-error behavior already documented in
  \passthrough{\lstinline!STATUS.md!} for
  \passthrough{\lstinline!divergence(gradient(psi))!} (same convergence
  order but a different leading constant). The direct stencil sidesteps
  that drift and lands the roadmap acceptance bar at the published
  constant.
\item
  \textbf{Six-term sum} at each grid point: three single-axis 4th
  derivatives (\(\partial_x^4, \partial_y^4, \partial_z^4\)) using
  \passthrough{\lstinline!stencil::FourthDerivative<Order>!}, plus three
  mixed (2, 2) cross-terms
  (\(\partial_x^2\partial_y^2, \partial_x^2\partial_z^2,
  \partial_y^2\partial_z^2\)) using the outer product of two
  \passthrough{\lstinline!stencil::Coefficients<Order>!} tables along
  the relevant axis pair. Per-axis scaling: \(1/h_a^4\) for the
  single-axis terms, \(1/(h_a^2 h_b^2)\) for the cross-terms. Per-point
  reads: order 2 → 27 (3 × 5 single-axis

  \begin{itemize}
  \tightlist
  \item
    3 × 3 × 3 cross-terms with overlap); order 4 → 21 + 75 = 96 (rough).
  \end{itemize}
\item
  \textbf{\passthrough{\lstinline!d4<Order>(psi)!}} is a 1-line alias
  forwarding to \passthrough{\lstinline!biharmonic<Order>!}.
\end{enumerate}

\subsubsection{Test surface}\label{test-surface}

\href{../../../test/phase_8_test.cpp}{\passthrough{\lstinline!test/phase\_8\_test.cpp!}}
holds:

\begin{itemize}
\tightlist
\item
  \textbf{Weight-table audits} ---
  \passthrough{\lstinline!EXPECT\_DOUBLE\_EQ!} on every weight in
  \passthrough{\lstinline!ThirdDerivative<2>!},
  \passthrough{\lstinline!ThirdDerivative<4>!},
  \passthrough{\lstinline!FourthDerivative<2>!},
  \passthrough{\lstinline!FourthDerivative<4>!} against the values
  derived in §2.
\item
  \textbf{62 convergence sweeps} generated by the
  \passthrough{\lstinline!EXPECT\_PARTIAL\_SLOPE\_RANGE(NAME, NX, NY, NZ, ORDER)!}
  macro --- one per (named function, Order) pair across the 31 partials,
  exercising the user-facing API explicitly so the
  \passthrough{\lstinline!<Order>!} template binding is part of the
  test.
\item
  \textbf{2 biharmonic convergence sweeps} + \textbf{4 explicit
  \(\nabla^4\psi = |\mathbf{k}|^4\psi\) acceptance tests} at \(N = 64\).
\item
  \textbf{2 alias-parity tests} that verify
  \passthrough{\lstinline!d4<Order>(psi)!} returns the same data as
  \passthrough{\lstinline!biharmonic<Order>(psi)!} element-wise.
\end{itemize}

\subsection{4. Design decisions}\label{design-decisions}

These distill the choices recorded in
\passthrough{\lstinline!specs/2026-05-02-phase-8-higher-order-derivatives/requirements.md!}.

\begin{itemize}
\item
  \textbf{Scope expansion to all 31 unique partials.} The roadmap
  originally scoped Phase 8 to ``rank-2 mixed partials + biharmonic.''
  The user expanded scope during the spec round to also ship single-axis
  3rd \& 4th partials and all mixed 3rd / 4th partials, on the argument
  that the multi-index machinery is the same code regardless of which
  subset is exposed and shipping the full set now avoids a follow-up
  phase. Roadmap was edited in the same branch to match.
\item
  \textbf{\passthrough{\lstinline!d<rank>d<axis-with-exponent>...!}
  naming convention with lexicographic axis order.} The alternative ---
  a unified \passthrough{\lstinline!derivative<Rank, Axes...>!} template
  --- was rejected because at the call site the named form is more
  self-documenting and the function-list explosion is bounded (31
  functions) and one-time. The lexicographic-ordering rule collapses
  operator-equivalent permutations (\passthrough{\lstinline!d3dx2dy!}
  and \passthrough{\lstinline!d3dydx2!} represent the same operator on
  smooth fields; only the lexicographic form is named) and keeps the
  public surface deterministic.
\item
  \textbf{Direct stencils for biharmonic, not
  \passthrough{\lstinline!laplacian∘laplacian!}.} The composition would
  converge at the right order but with a different leading-error
  constant --- the same stride-2 phenomenon already documented for
  \passthrough{\lstinline!divergence(gradient(psi))!} in Phase 2's
  STATUS entry. The roadmap acceptance bar (\(\nabla^4\psi\) recovers
  \(|\mathbf{k}|^4\) on \(\sin(\mathbf{k}\cdot\mathbf{x})\)) is cleanest
  with the direct six-term stencil.
\item
  \textbf{Sibling templates
  \passthrough{\lstinline!ThirdDerivative<Order>!} and
  \passthrough{\lstinline!FourthDerivative<Order>!}, not a unified
  \passthrough{\lstinline!Derivative<Rank, Order>!}.} Same call as Phase
  2 made for \passthrough{\lstinline!FirstDerivative<Order>!} (sibling
  to Phase 1's \passthrough{\lstinline!Coefficients!}). Renaming the
  existing tables would be a public-API break; adding new siblings is
  additive. The unification meta-decision was considered for the second
  time in Phase 8 and again deferred --- without a use case where
  \passthrough{\lstinline!Rank!} would be a meta-template parameter, the
  redesign cost exceeds its benefit.
\item
  \textbf{Hard-coded weight tables, no constexpr Fornberg generator.}
  Phase 8 ships eight new specializations totaling \textasciitilde30 LOC
  of weights. Fornberg becomes worth its weight when a phase needs
  orders \(\geq 6\) or many ranks at once; that decision moves to Phase
  11 (up-to-4th-order gradient) at the earliest. Audit of the hard-coded
  values is delegated to \passthrough{\lstinline!EXPECT\_DOUBLE\_EQ!}
  against published Fornberg-derived references in the test suite.
\item
  \textbf{Convergence-sweep grid sizes
  \passthrough{\lstinline!N ∈ \{16, 32, 64\}!}.} Same triplet as Phases
  1, 2, 7 --- keeps the slope window familiar and CI duration
  manageable. The order-4 rank-(2,2) cross stencil has radius 3 per
  axis; at \(N = 16\) the halo-to-grid ratio is \(3/16 = 19\%\), safe
  under \passthrough{\lstinline!IndexPolicy::Periodic!}.
\end{itemize}

\subsection{5. References}\label{references}

External:

\begin{itemize}
\tightlist
\item
  Fornberg, B. (1988). \emph{Generation of Finite Difference Formulas on
  Arbitrarily Spaced Grids}. \textbf{Mathematics of Computation},
  51(184), 699--706.
  \url{https://doi.org/10.1090/S0025-5718-1988-0935077-0}. The generic
  algorithm for deriving central-difference weights of arbitrary
  derivative rank and accuracy order; used as the \emph{theoretical}
  derivation tool for the four new tables in §2 even though the project
  hard-codes the resulting values.
\item
  LeVeque, R. J. (2007). \emph{Finite Difference Methods for Ordinary
  and Partial Differential Equations: Steady-State and Time-Dependent
  Problems}. SIAM. ISBN 978-0-89871-629-0.
  \url{https://doi.org/10.1137/1.9780898717839}. Chapter 1, Section 1.3
  (``Higher order accuracy'') gives the standard 5-point
  fourth-difference stencil and the truncation analysis that yields the
  order-2 fourth-derivative weights \(\{1, -4, 6, -4, 1\}\).
\item
  NIST Digital Library of Mathematical Functions, §1.10 (``Differential
  Operators''). \url{https://dlmf.nist.gov/1.10}. Permanent reference
  for the formal definition of \(\nabla^4\) as \((\nabla^2)^2\) and the
  resulting six-term expansion in 3D used in
  \passthrough{\lstinline!Biharmonic.hpp!}'s implementation.
\end{itemize}

Internal cross-references (do not satisfy the ``external'' minimum):

\begin{itemize}
\tightlist
\item
  \href{phase-7-higher-order-stencils.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-7-higher-order-stencils.md!}}
  --- the \passthrough{\lstinline!Coefficients<4>!} and
  \passthrough{\lstinline!FirstDerivative<4>!} weight tables derived in
  Phase 7 are reused unchanged by Phase 8's
  \passthrough{\lstinline!mixedDerivative!} helper for any
  \passthrough{\lstinline!(N, Order)!} with
  \passthrough{\lstinline!N ∈ \{1, 2\}!}.
\item
  \href{phase-2-gradient-divergence.md}{\passthrough{\lstinline!docs/developer-note/notes/phase-2-gradient-divergence.md!}}
  --- for the documented stride-2 leading-error behavior of
  \passthrough{\lstinline!divergence(gradient(psi))!} that motivates the
  direct-stencil biharmonic implementation in §3.
\end{itemize}
